---
title: "Creating a model tree diagram"
author: "Kyle Barrett"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model-tree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
is_pdf <- knitr::is_latex_output()
if(isTRUE(is_pdf)){
  # takes screenshot of rendered plots and converts it into a static image
  # only needed if this vignette is rendered to PDF
  library(webshot)
  webshot::install_phantomjs() %>% suppressMessages()
}
```



```{r load packages, results = 'hide', message=FALSE, warning=FALSE}
# NOTE: if running chunks interactively we need to load the package first
#   because renv isolation prevents us from finding an bbr installation
if (interactive()) {
  devtools::load_all()
}
devtools::load_all()
library(glue)
library(bbr)
library(dplyr)
library(purrr)
```

```{r set up and define cleanup, include = FALSE}
MODEL_DIR <- system.file("model", "nonmem", "basic", package = "bbr", mustWork = TRUE)
source("../tests/testthat/helpers-create-example-model.R")

# set bbi execution path
if (Sys.which("bbi") == "") {
  options('bbr.bbi_exe_path' = read_bbi_path())
}

# Get bbi version: we cant build this vignette when bbi is less than 3.2.0
# since model_summary() is required for the tooltip
eval_model_tree <- bbr:::test_bbi_version(read_bbi_path(), "3.2.0")

# Create fake runs
copy_output_dir <- function(orig_mod, new_mod) {
  new_dir_path <- new_mod[["absolute_model_path"]]
  fs::dir_copy(orig_mod[["absolute_model_path"]], new_dir_path)
  
  # replace file names with new model ID
  orig_mod_id <- get_model_id(orig_mod)
  new_mod_id <- basename(new_dir_path)
  purrr::walk(fs::dir_ls(new_dir_path), ~ {
    if (stringr::str_detect(basename(.x), glue::glue("^{orig_mod_id}"))) {
      mod_name <- stringr::str_replace(basename(.x), glue::glue("^{orig_mod_id}"), new_mod_id)
      fs::file_move(.x, file.path(dirname(.x), mod_name))
    }
  })
}

# Create example working directory
create_tree_models <- function(MODEL_DIR){
  MOD1_PATH <- file.path(MODEL_DIR, 1)
  tags <- c("new tag 1", "new tag 2")
  # copy models before creating model tree
  mod1 <- read_model(MOD1_PATH)
  mod2 <- copy_model_from(mod1, "2", .overwrite = TRUE, .add_tags = tags) %>% 
    update_model_id() %>% add_star()
  mod3 <- copy_model_from(mod2, "3", .overwrite = TRUE, .add_tags = "DV: nmol") %>% 
    update_model_id()
  mod4 <- copy_model_from(mod2, "4", .overwrite = TRUE) %>% update_model_id()
  mod5 <- copy_model_from(mod4, "5", .overwrite = TRUE) %>% update_model_id()
  
  # Multiple based_on
  mod6 <- copy_model_from(
    mod5, "6", .overwrite = TRUE, .based_on_additional = c("1", "3"),
    .star = TRUE, .description = "final model"
  ) %>% update_model_id() %>% suppressWarnings()
  mod7 <- copy_model_from(
    mod6, "7", .overwrite = TRUE, .based_on_additional = c("2")
  ) %>% update_model_id() %>% suppressWarnings()
  
  # Fake runs
  copy_output_dir(mod1, mod2)
  copy_output_dir(mod1, mod3)
  copy_output_dir(mod1, mod4)
  copy_output_dir(mod1, mod5)
  copy_output_dir(mod1, mod6)
  
  # Fake Bootstrap run
  boot_run <- make_fake_boot(mod6, n = 10)
  
  # Fake simulation (creates a new model)
  mod_sim <- make_fake_sim(mod3, n = 10, mod_id = "8")
  
  # Return mods to delete (dont return mod1)
  return(list(mod2, mod3, mod4, mod5, mod6, mod7, boot_run, mod_sim))
}

# delete old files
cleanup <- function(mods) {
  delete_models(mods, .tags = NULL, .force = TRUE) %>% 
    suppressMessages()
}
```


# Create a model tree diagram

While the `run_log()`, `summary_log()`, and `config_log()` are helpful for summarizing the model development process, `model_tree()` provides a convenient way to visualize and track any of these tabulated parameters. You can create a tree diagram using a modeling directory or run log.
```{r fake model dir, eval = FALSE}
MODEL_DIR <- "../nonmem"
```

```{r, echo=FALSE, eval=eval_model_tree, message=FALSE, warning=FALSE}
mods <- create_tree_models(MODEL_DIR)
```


By default, models will be colored by their run number, and basic summary statistics will display as a tooltip when hovered over.
```{r, eval=FALSE}
model_tree(MODEL_DIR)
```

```{r, echo=FALSE, eval=eval_model_tree}
# you dont need to specify the width when rendering in Rstudio viewer, but it makes a difference when rendered in Rmarkdown 
# - relying on auto-sizing leads to a plot with a small width
model_tree(MODEL_DIR, width = 800)
```

If coloring by a logical column, `FALSE` and `TRUE` values will correspond to white and red coloring respectively. Numeric or character columns will be colored as a gradient. `NA` values will appear grey regardless of the column type.
```{r, eval=FALSE}
model_tree(MODEL_DIR, color_by = "star")
```

```{r, echo=FALSE, eval=eval_model_tree}
model_tree(MODEL_DIR, color_by = "star", width = 800)
```


Specific columns can be added to the tooltip via the `include_info` argument. Though you are limited to the default `run_log()` columns when passing a model directory, you can pass any available columns when passing a run log  dataframe (must inherit the class `bbi_log_df`). The examples below to illustrate cases where you may want to do that.
```{r run_log} 
log_df <- run_log(MODEL_DIR)
log_df
```

In this example we define a new column, `out_of_date`, to denote whether the model or data has changed since the last run. We can color by this new column to determine if any of the models need to be re-run:
```{r, eval=FALSE}
run_log(MODEL_DIR) %>%
  add_config() %>%
  dplyr::mutate(out_of_date = model_has_changed | data_has_changed) %>%
  model_tree(
    include_info = c("model_has_changed", "data_has_changed", "nm_version"),
    color_by = "out_of_date"
  )
```

```{r, echo=FALSE, eval=eval_model_tree}
# Since using fake model runs (appearing out of date), set most to FALSE
log_df <- run_log(MODEL_DIR) %>% add_config() %>% suppressWarnings()
log_df$model_has_changed[2:7] <- FALSE

log_df %>% 
  dplyr::mutate(out_of_date = model_has_changed | data_has_changed) %>%
  model_tree(
    include_info = c("model_has_changed", "data_has_changed", "nm_version"),
    color_by = "out_of_date",
    width = 800
  )
```

The model tree can also be helpful for quickly determine if any heuristics were found during any model submissions, as well as displaying specific model summary output in the tooltip. 
```{r, eval=FALSE}
run_log(MODEL_DIR) %>%
  add_summary() %>%
  model_tree(
    include_info = c("tags", "param_count", "eta_pval_significant"),
    color_by = "any_heuristics"
  )
```

```{r, echo=FALSE, eval=eval_model_tree}
# Since using fake models (all inheriting the same issues), set some to FALSE
log_df <- run_log(MODEL_DIR) %>% add_summary()
log_df$any_heuristics[c(5, 7, 9)] <- FALSE
model_tree(
  log_df,
  include_info = c("tags", "param_count", "eta_pval_significant"),
  color_by = "any_heuristics",
  width = 800
)
```

```{r cleanup bottom, include=FALSE, eval=eval_model_tree}
# delete old files
cleanup(mods)
```

