---
title: "rbabylon Parameter Labels"
author: "Seth Green"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{parameter-labels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

  ```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates extracting parameter estimates and labels into a table that can be used for diagnostics, or to generate reports. *Note, this vignette exclusively deals with NONMEM models.*

If you are new to `rbabylon`, the ["Getting Started with rbabylon"](getting-started.html) vignette will take you through some basic scenarios for modeling with NONMEM using `rbabylon`, introducing you to its standard workflow and functionality.

## Setup

There is some initial set up necessary for using `rbabylon`. Please refer to the ["Getting Started"](getting-started.html) vignette, mentioned above, if you have not done this yet. Once this is done, load the library and set your modeling directory.

```{r set_model_directory}
library(rbabylon)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(readr))

set_model_directory("../inst/param_examples")
```

```{r setup, include = FALSE}
MODEL_DIR <- "../inst/param_examples"
source("../tests/testthat/data/test-param-labels-ref.R")
REF_OMEGA <- PARAM_BLOCK_REF$PEX_BLOCK32S$ctl

# set babylon execution path
if (Sys.which("bbi") == "") {
  options('rbabylon.bbi_exe_path' = read_bbi_path())
}
```


# Parameter labels

## Create Model Object
Our modeling directory contains a control stream and output directory, because the model has already been run.

```{r first model}
mod1 <- read_model("510")
class(mod1)
```

The model object you have just created can now be passed to the post-processing functions to create your tables.

## Extract Parameter labels

Currently, the `param_labels()` function parses labels from the comments in the control stream. Here is the relevant section of our example control stream.

```{r read ctl 1}
mod1 %>% 
  get_model_path() %>%                          # get control stream file path
  read_lines(skip = 15, n_max = 18) %>%  # read in only parameter section
  paste(collapse = "\n") %>%
  cat()
```

This control stream is parsed into the tibble below, following the syntax defined in the "Details" section of the `?param_labels` documentation.

```{r summary 1}
label_df <- mod1 %>% 
  param_labels() %>% 
  apply_indices(.omega = block(2))
label_df
```

Note, there are some subtleties to the `apply_indices()` function that will be addressed in the next section.

## Add Parameter Estimates
The user can also extract parameter estimates using the `model_summary()` and `param_estimates()` functions.

```{r model_summary}
param_df <- mod1 %>% 
  model_summary() %>% 
  param_estimates()
param_df
```

These two tibbles can be joined together to create a table for including in reports.

```{r param_estimates}
report_df <- inner_join(
  label_df %>% select(-param_type),
  param_df %>% select(names, estimate, stderr),
  by = "names"
)
report_df
```

# Applying indices

Because there are numerous ways of specifying the diagonal and off-diagonal elements of an `$OMEGA` or `$SIGMA` block in a control stream, automatically parsing the structure of these blocks can be brittle and error prone. For this reason, indices are *not* automatically added to the output of the `param_labels()` function and are instead added with the  `apply_indices()` function.

## Default behavior: All Diagonal
By default `apply_indices()` assumes that *all* `$OMEGA` and `$SIGMA` elements are diagonal. If this is the case, you do not need to pass anything to `.omega` or `.sigma` arguments described below. However, be careful that you do not accidentally overlook this because your indices will be incorrectly returned as simply `(1,1)`, `(2,2)`, `(3,3)`, etc.

## Specifying Block Structure
Block structure is specified with two arguments, `.omega` and `.sigma` which each take a logical vector. Each element in this vector corresponds to a parameter specified in the control stream and denotes whether that element is a diagonal in the relevant matrix.

For example, an `$OMEGA BLOCK(3)` would be represented with the vector `.omega = c(TRUE, FALSE, TRUE, FALSE, FALSE, TRUE)` because the first, third, and sixth elements are the diagonals and the others represent the off-diagonals. However, the user doesn't need to type this explicitly because `rbabylon` has a helper function `block()` that generates these vectors.

```{r block demo}
cat(paste("block(1): ", paste(block(1), collapse = ", "), "\n"))
cat(paste("block(2): ", paste(block(2), collapse = ", "), "\n"))
cat(paste("block(3): ", paste(block(3), collapse = ", "), "\n"))
cat(paste("block(4): ", paste(block(4), collapse = ", "), "\n"))
```

Notice the use of `.omega = block(2)` in the example from the previous section.

## Mixed structure

More complicated block structures can be represented by concatenating these logical vectors together. For example, the following omega block:

```{r ref omega}
cat(REF_OMEGA)
```

```{r parse ref}
REF_OMEGA %>%
  param_labels() %>%
  apply_indices(
      .omega = c(block(3), block(2), block(1), block(1))
    )
```

It is equally valid to replace any of these with logical vectors as well. For instance, instead of `c(..., block(1), block(1))` it may be easier to write `c(..., rep(TRUE, 2))`, especially if there are a number of `BLOCK(1)` lines in a row in the control stream.

## SAME blocks
It is also worth noting here that blocks which use `SAME` (for example, `$OMEGA BLOCK(1) SAME` above) will inherit the label from the block to which `SAME` refers. This is true even if there are multiple `SAME` blocks in a row.
